---
title: "Mean, Median, Standard Deviation, Correlation, Minimum and Maximum"
author: Yuk Tung Liu
output: 
  html_document:
    theme: readable
    toc: true
fontsize: 18pt
---
<style type="text/css">

body, td {
   font-size: 18px;
}
code.r{
  font-size: 18px;
}
pre {
  font-size: 18px
}
</style>

<!-- script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "all" } }
});
</script -->

```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(comment = NA)
```

Here we introduce several useful functions in R: `mean()`, `median()`, `max()`, `min()`, `sort()`, `var()`, `sd()`, `cor`, `which.min` and `which.max`.

## Mean, Median, Maximum, Minimum and Sorting

Let's first create a numeric vector: 
```{r}
x <- c(1.4, 5.66, 7.13, 9.21)
```
The `max()` and `min()` functions return the maximum and minimum of a vector: 
```{r}
max(x)
```
```{r}
min(x)
```
The `mean()` and `median()` functions return the mean and median: 
```{r}
mean(x)
```
```{r}
median(x)
```
The `sort()` function sorts the vector in increasing order:
```{r}
sort(x)
```
There are other optional parameters we can set in these functions. Use, e.g., `?sort` in the R console to pull up a help page. For example, we can sort the vector in decreasing order using 
```{r}
sort(x, decreasing=TRUE)
```

## Variance and Standard Deviation

The `var()` and `sd()` functions calculate the variance and standard deviation of a vector. Note that they are defined as 
$$ {\rm var}(X)= \frac{1}{N-1}\sum_{i=1}^N (X_i -\mu_X)^2 \ \ \ , \ \ \ {\rm sd}(X)=\sqrt{{\rm var}(X)} ,$$
where $N$ is the number of elements in $X$ and $\mu_X$ is the mean of $X$. These are called the *sample variance* and *sample standard deviation*. They are used to estimated the population variance and population standard deviation based on the random sample. Recall that as you have learned in Stat 100, the population variance $\sigma^2$ is defined as 
$$ \sigma^2 = \frac{1}{N}\sum_{i=1}^N (X_i -\mu_X)^2$$
So you have to be careful in using the `var()` and `sd()` functions in R. For example, if we want to calculate the population standard deviation, defined as the square root of  $\sigma^2$ above, for the `x` vector, we need to use the expression 
```{r}
stdx <- sd(x) * sqrt((length(x)-1)/length(x))
```
Here `length(x)` returns the number of elements in the vector `x`: 
```{r}
length(x)
```
The factor `sqrt((length(x)-1)/length(x))`, or $\sqrt{(N-1)/N}$, takes into account the $1/N$ and $1/(N-1)$ differences between the `sd()` function and the expression for $\sigma$. Another way of calculating $\sigma$ is to use the fact that $\sigma$ is the square root of the mean of $(X-\mu_X)^2$:
```{r}
stdx2 <- sqrt( mean( (x-mean(x))^2 ) )
```
We can confirm that these two expressions give the same result: 
```{r}
c(stdx, stdx2, stdx-stdx2)
```

## Correlation

Now that we have mean and standard deviation, we can calculate the Z-score, defined as 
$$ Z_i = \frac{X_i-\mu_X}{\sigma}$$
This can be done easily with R's vectorized operation:
```{r}
Zx <- (x-mean(x))/stdx
Zx
```
Consider another numeric vector `y` of length 4:
```{r}
y <- c(1.53, -3.45, 6.7, 4.63)
```
We can easily convert it to the Z score as well:
```{r}
stdy <- sd(y) * sqrt((length(y)-1)/length(y))
Zy <- (y-mean(y))/stdy
Zy
```
Recall that the correlation coefficient $r$ of two sets of variables X and Y is defined as the mean of the product their Z scores:
$$ r = \frac{1}{N}\sum_{i=1}^N Z_{xi} Z_{yi} $$
The correlation between the vectors `x` and `y` can be calculated using 
```{r}
mean(Zx*Zy)
```
Since the correlation coefficient is an important concept in statistics, R already has a built-in function, `cor()`, to compute it directly: 
```{r}
cor(x,y)
```
To demonstrate another use of the `cor()` function, we create a third vector: 
```{r}
z <- c(2.39,3.19,8.31,-4.67)
```
With 3 sets of data x, y, z, we can calculate the correlation coefficients between all pairs of the 3 variables: 
```{r}
cor(x,y)
cor(x,z)
cor(y,z)
```
There is a simpler way of calculating all 3 correlations by combining the `cbind()` function and the `cor()` function. 

The `cbind()` function is mentioned in Section 5.9 of the textbook. It combines vectors into a matrix whose columns are the input vectors. For example, 
```{r}
cbind(x,y,z)
```
creates a 4&times;3 matrix. The first column is `x`, second column is `y` and third column is `z`.

<p id="corMatrix">The `cor()` function accepts a matrix argument. In `cor(m)`, if `m` is a $p\times q$ matrix, `cor(m)` returns a $q \times q$ matrix whose $(i,j)$ element is the correlation coefficient between the $i$th column vector and $j$th column vector of `m`. For example, in </p>
```{r}
cor(cbind(x,y,z))
```
the (1,1) element is `cor(x,x)`, which is 1; the (1,2) element is `cor(x,y)`; the (1,3) element is `cor(x,z)`; the (2,1) element is `cor(y,x)`, which is equal to `cor(x,y)`; the (2,2) element is `cor(y,y)`, which is 1; the (2,3) element is `cor(y,z)`, and so on. The returned $3\times 3$ matrix is called the *correlation matrix* for the variables $x$, $y$, and $z$.

##Position of Maximum and Minimum

The `max()` and `min()` functions return the maximum and minimum value in a vector, but they don't tell us *where* the maximum and minimum occurs. If we want that information, we can use the `which.min()` and `which.max()` functions, which return the index where the first maximum and minimum occurs. As an example, consider 5 students in Stat 100 with the following heights and weights:
```{r}
height <- c(63,71,78,72,73)
weight <- c(120,167,190,165,290)
```
In this data set, height is measured in inches and weight is measued in pounds. The height of the tallest student is 
```{r}
max(height)
```
which is `r max(height)` inches. Suppose we want to know the weight of this tallest student. We can first find out the index of the maximum in the `height` vector and then look at the value of the `weight` vector at that index:
```{r}
i <- which.max(height)
i
weight[i]
```
The command `which.max(height)` returns `r i`, indicating that the maximum value in `height` is the `r i`rd element. The weight of this tallest student is `weight[3]` or `r weight[i]` pounds. Similarly, if we want to find the weight of the shortest student, we will use `i <- which.min(height)` to find the index of the minimum and then `weight[i]` to find the corresponding value of weight. 

Note that `which.min()`/`which.max()` returns the index of the *first* minimum/maximum in case there are ties. For example, if we have the following vector `x`.
```{r}
x <- c(1,5,10,10,5,1)
```
Then `which.max(x)` returns 3 since the third element is the first index where the maximum occurs. Similarly, `which.min(x)` returns 1. In many applications, maximum/minimum only occurs once and so `which.min()` and `which.max()` are sufficient. In the case of ties and we want to find all the indices, we will use the function `which()` combined with `min()` or `max()`. For example, to find all the indices of the minimum in `x`, we can use the command
```{r}
which(x==min(x))
```
You will learn the function `which()` in lesson 8 of swirl.

<br />
<br />
<br />
