---
title: "Using R as a Calculator"
author: Yuk Tung Liu
output: 
  html_document:
    theme: readable
fontsize: 18pt
---
<!-- Written by Yuk Tung Liu -->
<style type="text/css">

body, td {
   font-size: 18px;
}
code.r{
  font-size: 18px;
}
pre {
  font-size: 18px
}
</style>

<!-- script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "all" } }
});
</script -->

```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(comment = NA)
```

[//]: require files - harmonic_mean_solution.html, geometric_mean_solution.html

You should have completed lessons 1 and 3 of R programming in swirl before reading this article.

One thing you notice when opening an R console is that you can use it as a calculator. In addition to the basic arithmetic operations addition (+), subtraction (-), multiplication (*) and division (/), R has built-in standard mathematical functions. The following is a short list of standard mathematical functions.

Function                    Name                                                  Example 
---------------------- -----------------------------------------------  ----------------------------------
abs                      absolute value                                        abs(3-6)   = 3  
sqrt                     square root                                          sqrt(16)  = 4 
^                       exponentiation                                      3^10 = $3^{10}$  = 59049
exp                      exponential function                                  exp(1.7) = $e^{1.7}$  = 5.473947 
log                      log function (base e)                              log(10)  = 2.302585
log10                    base 10 log ($\log_{10}$)                             log10(100)  = 2
pi                      mathematical constant $\pi$                         pi  = 3.141593
sin, cos, tan           trigonometric functions (argument in radians)      sin(pi/2)  = 1
asin, acos, atan        inverse trigonometric functions                     acos(1)  = 0
sinh, cosh, tanh        hyperbolic functions                                cosh(0) = 1
asinh, acosh, atanh     inverse hyperbolic functions                        atanh(tanh(12)) = 12
round(x,n)               round x to n decimal places                        round(pi,2)  = 3.14
floor                          rounds down                                   floor(14.7)  = 14
ceiling                       rounds up                                      ceiling(14.7)  = 15

There are also useful statistical functions that we will talk about later. The ability of using R as a calculator allows us to analyze data interactively, as we will demonstrate later in the course. One distinct advantage of R over a conventional scientific calculator is its ability to perform vectorized operations, as demonstrated below.

## Vectorized Operations

As explained in lesson 3 of R programming in swirl, we can generate a sequential list of integer vector using `:`. For example, `1:100` generates a vector of length 100 with values 1, 2, 3, ... 100. Here "vector" simply means an array of numbers/characters/objects of the same class. We can also store the integer vector to a variable using the assignment operator `<-`: 
```{r}
x <- 1:100
``` 
Many other programming languages use `=` as an assignment operator. In R, you can use `=` as an assignment operator too. For example, `x = 1:100` is equivalent to `x <- 1:100` in this context. However, as you will learn later that the `=` operator in R has other uses. Assignments can also be made in the other direction, using the obvious change in the assignment operator. For example, `1:100 -> x` is equivalent to `x <- 1:100`. 

Suppose now I type 
```{r}
x <- 2*x - 1
```
What happens is that the original integer vector (1, 2, 3, ..., 100) is replaced by (1, 3, 5, 7, ..., 199) as each element is multiplied by 2 and then subtracted by 1. You can confirm this by typing `x` to auto print its content:
```{r}
x
```
What happens if I type the following? 
```{r}
x <- c(1,-1)*x
```
As explained in lesson 1 of R programming in swirl, since `c(1,-1)` is a vector of length 2 and `x` is a vector of length 100, R "recycles" the `c(1,-1)` vector 50 times to carry out the multiplication. The result is that the first element of the original vector is multiplied by 1, the second element is multiplied by -1, the third is multiplied by 1 and so on. So the content of x becomes: 
```{r}
x
```
Now let's do the following: 
```{r}
x <- sum(1/x)
```
`sum()` is a built-in R function that returns the sum of a vector. Thus, `x` is now just a number (numeric vector of length = 1): 
```{r}
x
```
It is the result of the sum
$$ 1-\frac{1}{3}+\frac{1}{5}-\frac{1}{7}+\cdots + \frac{1}{199}=\sum_{n=1}^{100} \frac{(-1)^{n-1}}{2n-1} $$
The same calculation can be compressed into a one-line expression: 
```{r}
sum( c(1,-1)/(2*(1:100)-1) )
```
or this one-line expression
```{r}
sum(c(1,-1)/seq(1,199,2))
```
The `seq()` function is introduced in lesson 3 of the R programming in swirl. If you forget how it is used, type `?seq` in the R console to pull up a help page. 

We see that a lengthy calculation can be carried out by just a one-line expression. If you are not impressed yet, try this: 
```{r}
s <- 4*sum( c(1,-1)/(2*(1:1e6)-1) )
```
By typing the above single line expression, we have just told R to carry out a sum over one million terms and then multiply by 4! The number `1e6` means `1000000` (1 followed by 6 zeros, or $10^6$), which is one million. The value stored in the variable `s` is equal to 
$$4\left( 1-\frac{1}{3}+\frac{1}{5}-\frac{1}{7}+\cdots + \frac{1}{1999999}\right) = 4\sum_{n=1}^{10^6} \frac{(-1)^{n-1}}{2n-1} $$
The value is 
```{r}
s
```
If this number seems familiar to you, it's because it's close to the number $\pi= 3.141592653589...$. In R, the variable `pi` stores this number: 
```{r}
pi
```
By default, R displays floating-point numbers to 7 significant figures even though R uses 8 bytes to store a floating-point number, corresponding to about 16 significant figures. You can change this default by the command `options(digits=n)`, where `n` is the number of significant figures you want R to display. For example, 
```{r}
options(digits=15)
```
sets the default printout to 15 significant figures: 
```{r}
c(s,pi)
```
So we see that `s` and `pi` only agree in the first 6 digits. In fact, it is well-known by mathematicians that the infinite series  
$$ 4 \sum_{n=1}^{\infty} \frac{(-1)^{n-1}}{2n-1} $$
converges to $\pi$ but the convergence is very slow. 

Now you can tell your friends that you have just learned a new skill to quickly compute the sum of a long series. You show your friends that you can sum the series of two million terms 
$$ 1+\frac{1}{2^2}+\frac{1}{3^2}+\cdots + \frac{1}{(2\times 10^6)^2}$$
by just typing
```{r}
sum(1/(1:2e6)^2)
```
Be aware, though, that if one of your friends is knowledgeable in math, he/she will laugh and point out that 
$$ \sum_{n=1}^{\infty} \frac{1}{n^2}=\frac{\pi^2}{6} \ ,$$
which was first figured out by the famous mathematician Leonhard Euler in 1735. You will be amazed and type  
```{r}
pi^2/6
```
to confirm that your sum is indeed close to this number. 

<!-- 
<blockquote>
Exercise 1: The *harmonic mean* of positive real numbers $x_1, x_2, \cdots, x_n$ is defined as 
$$H = \frac{n}{\frac{1}{x_1}+\frac{1}{x_2}+\cdots + \frac{1}{x_n}}$$ 
Write an R expression to calculate the harmonic mean of the numbers 1, 2, 3, &middot;&middot;&middot;, 1000. 

<a href="harmonic_mean_solution.html" target="_blank">See solution</a>.
</blockquote>
-->

The `sum()` function returns the sum of a vector. The `prod()` function returns the product of a vector. For example, the vector `3:5` is an integer vector consisting of 3, 4, and 5, whereas `prod(3:5)` returns $3\times 4\times 5$ or 60:
```{r}
prod(3:5)
```
Like the `sum()` function, `prod()` can be useful in some problems. Let's consider the birthday problem encountered in Stat 100 to further illustrate its use. 

### The Birthday  Problem

What is the probability that at least two people share the same birthday in a set of $n$ randomly chosen people? This is the famous birthday problem. If you forget how to solve it, look up your Stat 100 note or visit <a href="http://mathforum.org/dr.math/faq/faq.birthdayprob.html" target="_blank">this website</a>. The answer is given by the expression (ignore the leap date)
$$P = 1 - \frac{365}{365}\cdot \frac{364}{365}\cdot \frac{363}{365}\cdots \frac{366-n}{365}$$
In a class of 100 students, the probability that at least two students share the same birthday is
$$
  P = 1 - \frac{365}{365}\cdot \frac{364}{365}\cdot \frac{363}{365}\cdots \frac{266}{365} 
$$
What is the numerical value of this $P$? Your Stat 100 instructor told you that it's very close to 1. Do you believe it? Have you checked the calculation? When I first read the birthday problem from a book, I was suspicious of the claim. I used a calculator to carry out the calculation and confirmed the result. It took me a few minutes to finish the calculation. With R, it can be done by the following one-line expression: 
```{r}
1-prod((365:266)/365)
```
We see that it is indeed very close to 1. Let's break down the expression to see why it gives the desired result. `365:266` is an integer vector containing 365, 364, 363, ..., 266. `(365:266)/365` divides each element in `365:266` by 365, so it's a vector containing 
$$ \frac{365}{365}, \ \ \ \ \  \frac{364}{365}, \ \ \ \ \  \frac{363}{365}, \ \ \ \ \  \cdots \ \ \ \ \  \frac{266}{365}$$ 
`prod((365:266)/365)` returns the product of the vector in `(365:266)/365`. 
That is, 
$$ {\rm prod((365:266)/365)} = \frac{365}{365}\cdot \frac{364}{365}\cdot \frac{363}{365}\cdots \frac{266}{365}$$ 
which is the probability that all of the 100 students have different birthdays. This is a very small number: 
```{r}
prod((365:266)/365)
```
Finally, `1-prod((365:266)/365)` is the probability that at least two students share the same birthday. Since `prod((365:266)/365)` is very small ($\approx 3 \times 10^{-7}$), `1-prod((365:266)/365)` $\approx 1$ as claimed.

Interestingly, R has a built-in function for the birthday problem. It is called the probability of coincidences. The function `pbirthday(n)` returns the probability that at least 2 people share the same birthday among $n$ randomly chosen people. For $n=100$, the function gives 
```{r}
pbirthday(100)
```
exactly the same value calculated above. `pbirthday()` has other optional parameters you can specify for generalized birthday problems. There is also an associated function `qbirthday()`. Type `?pbirthday` for more information. 

<!--
<blockquote>
Exercise 2: The *geometric mean* of positive numbers $x_1, x_2,\cdots, x_n$ is defined as 
$$G = (x_1 x_2 \cdots x_n)^{1/n}$$
Write an R expression to calculate the geometric mean of 1, 2, 3, &middot; &middot; &middot;, 100. Can you use the same expression to calculate the geometric mean of 1, 2, 3, &middot; &middot; &middot;, 1000? If not, how would you calculate it?

<a href="geometric_mean_solution.html" target="_blank">See solution</a>.
</blockquote>
-->

<br />
<br />
<br />